Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2017-05-26 paper.tex
...

\subsubsection{Case Study: Advancing Pages in the \emph{Story screen}}

%\begin{figure}[t]
%\centering
%\includegraphics[width=\columnwidth]{story-anim}
%\caption{The \emph{Story screen}.
%\label{fig.story}
%}
%\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{story}
\caption{Continuation mechanism and state machine for advancing pages in the \emph{Story screen}.
\label{fig.story}
}
\end{figure}

\begin{figure*}[t]
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
StoryScreenComponent::StoryScreenComponent (<...>) :
    <...>
{
    pages        = <...>; // vector with loaded pages
    current_page = pages.back(); // first loaded page
    displayed    = false; // if current is complete
    <...>
}

<...>   // draw page over time

void StoryScreenComponent::update (<...>) {
    <...>
    if (<all-words-appearing>) {
        displayed = true;
    }
}

void StoryScreenComponent::next_text() {
    if (!displayed) {
        displayed = true;
        <...>     // remove current page
    } else {
        pages.pop_back();
        if (!pages.empty()) { // next page
            current_page = pages.back();
            displayed    = false;
            <...>
        } else {
            <...> // terminates the story screen
        }
    }
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Story (void) -> bool do
    <...>
    event void next_text; // clicks in >>>

    { pages = <...>; } // same as in C++
    loop i in [0 <- {pages.size()}[ do
        par/or do
            watching next_text do
                <...> // advance text
            end
            await next_text;
        with
            <...> // redraw _pages[i]
        end
    end
end

.
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Advancing pages in the \emph{Story Screen}.
\label{lst.story}
}
\end{figure*}

The clickable \emph{blue dots} in the campaign world map transit to ambience
story screens%
\footnote{Story screen animation: \url{github.com/an000/p/blob/master/README.md#4} }.
A story is composed of multiple pages and, inside each page, the words of the
story appear incrementally over time.
A first click in the button \code{>>>} fast forwards the words to show the full 
page.
A second click advances to the next page, until the story terminates.
If the page completes before a click (due to the time elapsing), a first click 
advances to the next page.
%

In C++, the class \code{StoryScreenComponent} in Figure~\ref{lst.story}.a
implements the method \code{next\_text}, which is a callback for clicks in the
button \code{>>>}.
%
The variable \code{pages} (ln. 4--5, 24--26) is a vector holding each page, but
which also encodes \emph{continuations} for the story progress:
each call to \code{next\_text} that advances the story (ln. 23--32) removes the 
current page (ln. 24) and sets the next action to perform (i.e., ``display a
new page'') in the variable \code{current\_page} (ln. 26).
Figure~\ref{fig.story} illustrates the continuation mechanism to advance 
pages and also a state machine for fast forwarding words (inside the dashed
rectangle).
The state variable \code{displayed} (ln. 6,15,20,21,27) switches between the
behaviors ``advancing text'' and ``advancing pages'', which are both handled
intermixed inside the method \code{next\_text}.

The code in \CEU in Figure~\ref{lst.story}.b uses the internal event
\code{next\_text}, which is emitted from clicks in \code{>>>}.
%
The sequential navigation from page to page uses a loop in direct style
(ln. 6--15) instead of explicit state variables for the continuation mechanism
and state machine.
While the text advances in an inner loop (hidden in ln. 9), we watch the
\code{next\_text} event that fast forwards it.
The loop may also eventually terminate with the time elapsing normally.
This way, we do not need a variable (such as \code{displayed} in C++) to switch 
between the states ``advancing text'' and ``advancing pages''.
The \code{par/or} makes the page advance logic to execute in parallel with the
redrawing code (ln. 13).
Whenever the page advances, the redrawing code is automatically aborted
(due to the \code{or} modifier).
The \code{await next\_text} in sequence (ln. 11) is the condition to advance to
the next page.
%
Note that, unlike the implementation in C++, the ``advancing text'' behavior is
not intermixed with the ``advancing pages'' behavior, instead, it is
encapsulated inside the inner loop nested with a deeper indentation (ln. 9).

@@

The direct style of \CEU has some advantages in comparison to the 
continuation-passing style:
%
\begin{itemize}
\item It uses structured control flow (i.e., sequences and loops) instead of 
      explicit data structures (e.g., stacks) and continuation variables (e.g.
      indexes).
\item The activities in sequence are decoupled and do not hold references to
      one another. % TODO: esta no outro exemplo
\item A single parent class describes the flow between the activities in a 
      self-contained block of code. % TODO: esta no outro exemplo
\end{itemize}

>>>>

\subsubsection{ Transition to the \emph{Credits Screen} from the \emph{Story Screen} }

\begin{figure*}[t]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
StoryDot::StoryDot(const FileReader& reader) :
  credits(false), // do not display by default
{
  <...>
  reader.read("credits", credits); // from file
}

void StoryDot::on_click() {
  <...>
  push_screen(<StoryScreen>(<...>, credits));
  <...>
}

///

StoryScreenComp::StoryScreenComp (<...>) :
  credits(credits),
  <...>
{
  <...>
}

<...>   // draw and update page

void StoryScreenComp::next_text() {
  if (!displayed) {
    <...>
  } else {
    <...>
    if (!pages.empty()) {
      <...>
    } else {
      if (credits) {
        replace_screen(<Credits>(<...>));
      } else {
        pop_screen();
      }
    }
  }
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
loop do
  var int ret = await Worldmap();
  if ret=={STORY_MAP} or ret=={STORY_CREDITS} then
    <...>
    var bool is_click = await Story();
    if is_click and ret=={STORY_CREDITS} then
      <...>
      await Credits();
    end
  else
    <...>
  end
end

.
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Transition from \emph{Story} to \emph{Credits screen}.
\label{lst.story}
}
\end{figure*}

The campaign world map has clickable blue dots for introductory and ending
ambience stories in the two extremes of the map progress trail.
For introductory stories, the game returns to the world map after displaying 
the story pages.
For ending stories, the game also displays a \emph{Credits screen} before
returning to the world map.%
\footnote{Story screen animation: \url{github.com/an000/p/blob/master/README.md#4} }

In C++, the class \code{StoryDot} in Figure~\ref{lst.story}.a (ln. 1--12) reads
the level file (ln. 5) to check whether its an ending story and should, after
termination, display the \emph{Credits screen}.
%
The boolean variable \code{credits} is passed to the class
\code{StoryScreen} (ln. 10) and represents the screen continuation, i.e., what
to do after displaying the story.
The class \code{StoryScreen} (not shown) then forwards the continuation even
further to the auxiliary class \code{StoryScreenComp} (ln. 16--40).
%
When the method \code{next\_text} has no pages left to display (ln. 32--38),
it decides where to go next, depending on the continuation flag
\code{credits} (ln. 33--37).

In \CEU, the \code{loop} of Figure~\ref{lst.story}.b controls the flow between
the screens to display as a direct sequence of statements.
%
We first invoke the \code{Worldmap} (ln. 2), which exhibits the map and let
the player interact with it until a dot is clicked.
If the player selects a story dot (ln. 4--9), we invoke the \code{Story}
and await its termination (ln. 5).
Finally, we check the returned values (ln. 6) to perhaps display the
\code{Credits} screen (ln. 8).
The enclosing loop restores the \code{Worldmap} and repeats the process.

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{continuation}
\caption{ Continuation (C++) vs Direct (\CEU) Styles.
\label{fig.story}
}
\end{figure}

Figure~\ref{fig.story} illustrates the \emph{continuation-passing style} of
C++ and the \emph{direct style} of \CEU for screen transitions:

\begin{enumerate}
\item \code{Main Loop} $\longrightarrow$ \code{Worldmap}:
    \begin{itemize}
    \item C++ uses an explicit stack to push the \code{Worldmap} screen.
    \item \CEU invokes the \code{WorldMap} screen expecting a return value
          (ln. 2).
    \end{itemize}
\item \code{Worldmap} (\emph{blue dot click}) $\longrightarrow$ \code{Story}:
    \begin{itemize}
    \item C++ pushes the \code{Story} screen passing the continuation flag
          (ln. 10).
    \item \CEU stores the \code{Worldmap} return value and invokes the \code{Story} screen
          (ln. 2,5).
    \end{itemize}
\item \code{Story} $\longrightarrow$ \code{Credits}:
    \begin{itemize}
    \item C++ replaces the current \code{Story} screen with the \code{Credits}
          screen (ln. 34).
    \item \CEU invokes the \code{Credits} screen after the \code{await Story}
          returns (ln. 8).
    \end{itemize}
\item \code{Credits} $\longrightarrow$ \code{Worldmap}:
    \begin{itemize}
    \item C++ pops the \code{Credits} screen, going back to the \code{Worldmap}
          screen.
          \CEU uses an enclosing \code{loop} to restart the process (ln. 1--13).
    \end{itemize}
\end{enumerate}

In contrast with C++, the screens in \CEU are decoupled and only the
\code{Main Loop} touches them: the \code{Worldmap} has no references to
\code{Story}, which has no references to \code{Credits}.

@@

The direct style of \CEU has some advantages in comparison to the 
continuation-passing style:
%
\begin{itemize}
\item It uses structured control flow (i.e., sequences and loops) instead of 
      explicit data structures (e.g., stacks) and continuation variables (e.g.
      boolean flags).
\item The activities in sequence are decoupled and do not hold references to
      one another. % TODO: esta no outro exemplo
\item A single parent class describes the flow between the activities in a 
      self-contained block of code. % TODO: esta no outro exemplo
\end{itemize}
